---
title: "Dichotomic Voom Differential Gene Expression analysis (DGE)"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
 %\VignetteIndexEntry{Dichotomic Voom Differential Gene Expression analysis (DGE)}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
# Description
Computes the differential gene expression analysis. The input consists of two groups of
RNA-Seq data from tissues under different conditions: control vs. altered state (e.g., a
disease-affected group and a control group).

RNA data must follow a negative binomial distribution, which is typical for RNA-Seq analysis.

For RNA data following a normal distribution (such as in microarray analysis), see
`DichotomicDGE`.

Derived class of `DichotomicDGEAbstract`.


```{r setup, include=FALSE}
library(DrugRepurposing)
```
```{r}
library(DrugRepurposing)
disease_dge <- data.frame(
  gene_id = c(1, 2, 3, 9, 11, 15),
  t.value = c(-9.261609, -10.36986, 8.17305166, 5.60715925, -8.444215, 7.12957687)
)

drug_dge <- data.frame(
  gene_id = c(1, 2, 3, 10, 11, 13),
  t.value = c(-2.26559, 1.369012, 3.173466, -4.60136, -2.191482, 4.2035724)
)

binChenDiseaseDGEDrugConnectivityScore <- BinChenDiseaseDGEDrugConnectivityScore$new()
binChenDiseaseDGEDrugConnectivityScore$compute(
  disease_dge = disease_dge,
  drug_dge = drug_dge,
  drug_dge_t_value_column_name = "t.value",
  random_distribution_size = 10, # 10 is a poor value used only to speed up
                                 # computation; an appropriate value is 10^5
  compute_p_value = TRUE,
  signature_mapper_parameter = 4
)
```
