---
title: "Bin Chen Disease Signature Drug List Connectivity Score Sync Facade"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
 %\VignetteIndexEntry{Bin Chen Disease Signature Drug List Connectivity Score Sync Facade}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
# Description
Computes the connectivity score between a disease signature and a list of drug DGEs.

Facade for `BinChenDiseaseSignatureDrugListConnectivityScoreCore`. It creates an instance
of `BinChenDiseaseSignatureDrugListConnectivityScoreCore` using the parallel
implementations of `RandomConnectivityScoreDistributionAbstract` and
`BinChenConnectivityScoreListApplyerAbstract`.

The parallel implementations used are: `RandomConnectivityScoreDistributionParallel` and `BinChenConnectivityScoreListApplyerParallel`.

This class is identical to `BinChenDiseaseSignatureDrugListConnectivityScoreSyncFacade`, except for the computation type.

Due to R architecture, it works only on Unix-like systems.
```{r setup, include=FALSE}
library(DrugRepurposing)
options(warn = -1)
```
```{r}
library(DrugRepurposing)

disease_signature <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 30, 31, 32, 33),
  estimate = c(-9.261609, -10.36986, 8.17305166, 5.60715925, -8.444215, 7.12957687,
               -1.4257425, -7.456546363, 8.97589585, 3.2533666, 4.32524634, -5.45856889)
)

drugs <- data.frame(
  name = c("A-23187", "A-443644", "AG-490", "AG-494"),
  filename = c("A-23187", "A-443644", "AG-490", "AG-494")
)


A_23187_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 31),
  estimate = c(1.98564, -1.51579, 0.33099, -0.935718, -3.818529,
               1.848352, -4.332752, 3.591474, -0.345235423)
)

A_443644_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 33),
  estimate = c(-1.55774, -4.4155, 1.21015, 4.8705936, -0.1259465,
               -2.6003531, -0.3530122, -3.5426363, -3.252352)
)

AG_490_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 37),
  estimate = c(-1.35143, 0.58724, -1.31592, 0.22267776, -0.45726445,
               -2.80993289, -0.05550198, 1.76931022, 0.1423542)
)

AG_494_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 6158),
  estimate = c(-3.43357, -0.14896, 0.90733, 3.084164, -2.594041,
               4.304979, -4.944015, -2.075178, 1.4523542)
)

drugs_rna_data <- list()
drugs_rna_data[["A-23187"]] <- A_23187_drug
drugs_rna_data[["A-443644"]] <- A_443644_drug
drugs_rna_data[["AG-490"]] <- AG_490_drug
drugs_rna_data[["AG-494"]] <- AG_494_drug


EasyDrugSignatureLoader <- R6Class(
  "EasyDrugSignatureLoader",
  inherit = DrugSignatureLoaderAbstract,
  public = list(
    initialize = function(drugs_rna_data) {
      private$drugs_rna_data <- drugs_rna_data
    },
    load = function(drug_name) {
      return(private$drugs_rna_data[[drug_name]])
    }
  ),
  private = list(
    drugs_rna_data = NA
  )
)

easyDrugSignatureLoader <- EasyDrugSignatureLoader$new(drugs_rna_data)

binChenDiseaseSignatureDrugListConnectivityScoreSyncFacade <-
  BinChenDiseaseSignatureDrugListConnectivityScoreSyncFacade$new(easyDrugSignatureLoader)
binChenDiseaseSignatureDrugListConnectivityScoreSyncFacade$compute(
  disease_signature = disease_signature,
  drugs = drugs,
  n_drug_signatures_genes = 9,
  random_distribution_size = 10,  # 10 is a poor value used only to speed up
  # computation; an appropriate value is 10^5
  disease_name = "IPF"
)

```
