---
title: "Dichotomic Voom Differential Gene Expression analysis (DGE)"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
  %\VignetteIndexEntry{Dichotomic Voom Differential Gene Expression analysis (DGE)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Description
Computes the differential gene expression analysis. The input consists of two groups of
RNA-Seq data from tissues under different conditions: control vs. altered state (e.g., a
disease-affected group and a control group).

RNA data must follow a negative binomial distribution, which is typical for RNA-Seq analysis.

For RNA data following a normal distribution (such as in microarray analysis), see
`DichotomicDGE`.

Derived class of `DichotomicDGEAbstract`.


```{r setup, include=FALSE}
library(DrugRepurposing)
```
```{r}
library(DrugRepurposing)
rna_seq <- matrix(
  c(
    262, 281897, 463, 12301, 284, 2966, 4842, 9966,
    116, 337026, 261, 3277, 132, 2006, 3699, 12406,
    111, 111736, 193, 862, 88, 1371, 2536, 1531,
    162, 219471, 281, 5990, 280, 2241, 4527, 12941
  ),
  nrow = 8,
  ncol = 4,
  dimnames = list(
    c(1, 2, 3, 9, 10, 11, 12, 13),
    c("GSM1", "GSM2", "GSM3", "GSM4")
  )
)

sample_01_map <- "1X01"
# How the map is applied:
#           1      X      0      1
# GeneID   GSM1   GSM2   GSM3   GSM4
# 1        262    116    111    162
# ...
#
#
# GSM1, GSM4 = IPF; GSM3 = control; GSM2 = excluded

filter_by_protein_coding <- FALSE

dichotomicVoomDGE <- DichotomicVoomDGE$new()
dichotomicVoomDGE$compute(rna_seq, sample_01_map, filter_by_protein_coding)
```
