---
title: "Dichotomic Differential Gene Expression analysis (DGE)"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
  %\VignetteIndexEntry{Dichotomic Differential Gene Expression analysis (DGE)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Description
Computes the differential gene expression analysis. The input is composed of two groups of
RNA data from tissues under different conditions: control/altered state; for example, a
disease-affected group and a control group. RNA data must follow the "normal distribution",
usually data from microarray-like analysis. For RNA data following a negative binomial
distribution, like Illumina RNA-Seq, see `DichotomicVoomDGE`.


```{r setup, include=FALSE}
library(DrugRepurposing)
```
```{r}
library(DrugRepurposing)
rna_data <- matrix(c(
3.000, 2.322, 2.807, 3.000,
9.676, 8.910, 9.006, 8.960,
4.954, 4.248, 3.907, 4.392,
1.000, 0.000, 0.000, 0.000,
0.000, 0.000, 0.000, 0.000,
1.000, 1.000, 1.585, 1.000,
0.000, 0.000, 0.000, 0.000,
5.883, 6.170, 6.459, 6.781
), nrow = 8, byrow = TRUE)
colnames(rna_data) <- c('GSM1', 'GSM2', 'GSM3', 'GSM4') # experiment names
rownames(rna_data) <- c(1, 2, 3, 9, 10, 11, 12, 13) # gene ids


# Gene id  GSM1    GSM2    GSM3    GSM
# 1        3.000   2.322   2.807   300
# 2        9.676   8.910   9.006   8.960
# 3        4.954   4.248   3.907   4.392
# 9        1.000   0.000   0.000   0.000
# 10       0.000   0.000   0.000   0.000
# 11       1.000   1.000   1.585   1.000
# 12       0.000   0.000   0.000   0.000
# 13       5.883   6.170   6.459   6.781

# the first column represents the gene id not the table's first column

sample_01_map <- "1X01"
# 0 = test sample (altered state)
# 1 = control state
# X = excluded

# How the map is applied:
#          1       X       0       1
# GeneID   GSM1    GSM2    GSM3    GSM4
# 1        3.000   2.322   2.807   3.000
# ...
#
#
# GSM1, GSM4 = IPF; GSM3 = control; GSM2 = excluded

filter_by_protein_coding <- FALSE

logarithmGeneFilter <- LogarithmGeneFilter$new(1)
dichotomicDGE <- DichotomicDGE$new(logarithmGeneFilter)
dichotomicDGE$compute(rna_data, sample_01_map, filter_by_protein_coding)
```
