---
title: "Dichotomic Voom Drug Repurpose"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
  %\VignetteIndexEntry{Dichotomic Voom Drug Repurpose}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Description
Given a disease RNA-Seq analysis and a list of drug DGEs, `DichotomicVoomDrugRepurpose`
 computes the connectivity score between the disease and the drugs.

This class composes the classes `DichotomicDGE` and `BinChenDiseaseDGEDrugListConnectivityScore`
through the `DichotomicDrugRepurposeCore` class.

Disease RNA-Seq data must be expressed as counts.

```{r setup, include=FALSE}
library(DrugRepurposing)
```
```{r}
library(DrugRepurposing)

rna_seq <- matrix(
  c(
    262, 281897, 463, 12301, 284, 2966, 4842, 9966,
    116, 337026, 261, 3277, 132, 2006, 3699, 12406,
    111, 111736, 193, 862, 88, 1371, 2536, 1531,
    162, 219471, 281, 5990, 280, 2241, 4527, 12941
  ),
  nrow = 8,
  ncol = 4,
  dimnames = list(
    c(1, 2, 3, 9, 10, 11, 12, 13),
    c("GSM1", "GSM2", "GSM3", "GSM4")
  )
)

sample_01_map <- "0X10"
drugs <- data.frame(
  name = c("A-23187", "A-443644", "AG-490", "AG-494", "AG-957", "AM-404"),
  filename = c("drugs_dge/A-23187.Rds", "drugs_dge/A-443644.Rds",
               "drugs_dge/AG-490.Rds", "drugs_dge/AG-494.Rds",
               "drugs_dge/AG-957.Rds", "drugs_dge/AM-404.Rds")
)
drugs_genes <- c(1, 2, 3, 9, 10, 11, 12, 13, 6158)


drugs <- data.frame(
  name = c("A-23187", "A-443644", "AG-490", "AG-494"),
  filename = c("A-23187", "A-443644", "AG-490", "AG-494")
)

A_23187_drug <- data.frame(
  gene_id = c(1, 2, 3, 9, 10, 11, 12, 13, 6158),
  estimate = c(1.98564, -1.51579, 0.33099, -0.935718, -3.818529,
               1.848352, -4.332752, 3.591474,-0.345235423)
)

A_443644_drug <- data.frame(
  gene_id = c(1, 2, 3, 9, 10, 11, 12, 13, 6158),
  estimate = c(-1.55774, -4.4155, 1.21015, 4.8705936, -0.1259465,
               -2.6003531, -0.3530122, -3.5426363,-3.252352)
)

AG_490_drug <- data.frame(
  gene_id = c(1, 2, 3, 9, 10, 11, 12, 13, 6158),
  estimate = c(-1.35143, 0.58724, -1.31592, 0.22267776, -0.45726445,
               -2.80993289, -0.05550198, 1.76931022,0.1423542)
)

AG_494_drug <- data.frame(
  gene_id = c(1, 2, 3, 9, 10, 11, 12, 13, 6158),
  estimate = c(-3.43357, -0.14896, 0.90733, 3.084164, -2.594041,
               4.304979, -4.944015, -2.075178,1.4523542)
)

drugs_rna_data <- list()
drugs_rna_data[["A-23187"]] <- A_23187_drug
drugs_rna_data[["A-443644"]] <- A_443644_drug
drugs_rna_data[["AG-490"]] <- AG_490_drug
drugs_rna_data[["AG-494"]] <- AG_494_drug


EasyDrugSignatureLoader <- R6Class(
  "EasyDrugSignatureLoader",
  inherit = DrugSignatureLoaderAbstract,
  public = list(
    initialize = function(drugs_rna_data) {
      private$drugs_rna_data <- drugs_rna_data
    },
    load = function(drug_name) {
      return(private$drugs_rna_data[[drug_name]])
    }
  ),
  private = list(
    drugs_rna_data = NA
  )
)

meanThresholdGeneFilter <- MeanThresholdGeneFilter$new(1)
easyDrugSignatureLoader <- EasyDrugSignatureLoader$new(drugs_rna_data)

dichotomicVoomDrugRepurpose <- DichotomicVoomDrugRepurpose$new(
  meanThresholdGeneFilter,
  easyDrugSignatureLoader
)
dichotomicVoomDrugRepurpose$compute(
  rna_seq = rna_seq,
  sample_01_map = sample_01_map,
  drugs = drugs,
  drugs_genes = drugs_genes,
  random_distribution_size = 10^4,
  disease_name = "IPF",
  filter_by_protein_coding = FALSE,
  parallel_computation = FALSE,
  signature_mapper_parameter = 8
)

```
