---
title: "Drug repurposing system based on differential expression analysis of RNA expression profiles"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
  %\VignetteIndexEntry{Drug repurposing system based on differential expression analysis of RNA expression profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Description
  This package provides a set of tools for analyzing genetic data from disease and drug studies to identify potential new uses for existing drugs, in other words, it serves as a drug repurposing system.

  An interesting application involves using genetic data from currently untreatable diseases to generate a list of candidate drugs for their treatment. The obtained results must then be validated in the laboratory.

  The basic idea is to analyze the sequencing of four transcriptomes from: healthy sample, diseased sample, diseased sample treated with a control vehicle, and diseased sample treated with drugs, in order to identify potential connections between drugs and diseases.

  This technique, known as "transcriptome signature reversion," has already been proposed and validated in several studies (e.g., [1,2]).

  The software presented in this package is based on the publications:

  1. Validation of transcriptome signature reversion for drug repurposing in oncology - Karel K. M. Koudijs et al. - 2022.
  2. Reversal of cancer gene expression correlates with drug efficacy and reveals therapeutic targets – Bin Chen et al. - 2017.


The repository accompanying the first article by Karel K. M. Koudijs et al., available at [https://gitlab.com/k.k.m.koudijs/TSR-comprehensive-validation](https://gitlab.com/k.k.m.koudijs/TSR-comprehensive-validation),
was used as the starting point for the development of DrugRepurposing.

The core class for drug repurposing, `BinChenCMapScoreByDrugRank`, was extracted from the software accompanying the second article by Bin Chen et al., available at [https://github.com/Bin-Chen-Lab/RGES](https://github.com/Bin-Chen-Lab/RGES).

Alternatively to `BinChenCMapScoreByDrugRank` is possible to use `connectivityScore` from `PharmacoGx`.

The transcriptome reversion system provided in this package implements the same mathematical model described by Karel K. M. Koudijs et al., but with significantly more efficient computational procedures. Thanks to concurrent programming, optimized algorithms, and the use of the Julia programming language, the new software is up to 70 times faster than the original version developed by Koudijs. However, the support of parallel processing is fully supported only on unix like systems.

## Drug repurposing analysis
The implemented repurposing pipeline consists of four principal stages:

1.	**Disease DGE Generation**. DGE is computed between diseased and healthy samples using either a dichotomic or linear mixed model (LMM)-based method, depending on study design and available metadata.


2.	**Drug DGE Generation**. DGE is computed for samples exposed to drugs versus vehicle-treated controls. This step is repeated for each compound under evaluation.


3.	**Signature Generation**. The most differential expressed genes are selected using either:
* the top n genes ranked by absolute t-value (typically 50–150), or

* genes with adjusted p-value < 0.001 and absolute log fold change above a specified threshold (e.g., 1.5).

4.	**Connectivity Score Computation**. A connectivity score is calculated to quantify the degree of inversion between the disease and drug signatures. A negative score suggests potential therapeutic relevance, indicating that the drug may reverse the disease gene expression pattern.


## Connectivity Score
The connectivity score represents the correlation between the disease and the drug. The score ranges from -1 to +1.

A positive value indicates that the drug behaves similarly to the disease.

A negative value indicates that the drug behaves conversely to the disease.

A higher positive score implies stronger correlation, while a higher negative score
indicates greater anticorrelation. Drugs with strongly negative scores are prioritized as repurposing candidates.


## Connectivity Score Calculation
The connectivity score algorithm is based on the model proposed by Lamb et al. [3] and refined by Chen et al. [2], the latter being the recommended implementation due to its robustness and specificity.


## Differential Expression Models
The package provides two classes of DGE methods:
* Dichotomic Models for pairwise comparisons (e.g., disease vs. control).
* Linear Mixed Models (LMM) for analyses involving additional factors such as tissue origin.

Support is provided for data following both normal distributions (e.g., microarray) and negative binomial distributions (e.g., RNA-Seq), with RNA-seq data requiring a preprocessing step via the voom transformation.


## Differential Gene Expression (DGE) analysis
This package provides four types of functions to compute differential gene expression (DGE) for four different types of RNA data:

1. **RNA data counts following a "negative binomial distribution" (typically RNA-seq data) with a single condition "altered state/control state"**.\
These functions are based on the voom transformation and `lmFit` from the `limma` package.
They are referred to as Dichotomic voom DGE.

2. **RNA data counts following a "negative binomial distribution" (typically RNA-seq data)  with additional conditions beyond the base "altered state/control state"**.\
 For example, in a disease analysis, samples may also differ by tissue of origin in addition to the "altered state/control state" condition.
These functions are based on the voom transformation and Linear Mixed Models (LMM).
The Linear Mixed Model is computed using one of the following functions: `lmer` from `lme4`, `dream` from `variancePartition`, or `fit` from Julia programming language.
They are referred to as LMM voom DGE.

3. **RNA data log-counts following a "normal distribution" (typically microarray data) with a single condition "altered state/control state"**.\
These functions are based on `lmFit` from the limma package.
They are referred to as Dichotomic DGE.

4. **RNA data log-counts following a "normal distribution" (typically microarray data) with additional conditions beyond the base "altered state/control state"**.\
For example, in a disease analysis, samples may differ by tissue of origin in addition to the "altered/control condition".
These functions are based on Linear Mixed Models (LMM).
The Linear Mixed Model is computed using one of the following functions: `lmer` from `lme4`, `dream` from `variancePartition`, or `fit` from Julia programming language.
They are referred to as LMM DGE.

For further information about voom, see:
*voom: Precision weights unlock linear model analysis tools for RNA-seq read counts - Charity W Law, Yunshun Chen, Wei Shi, Gordon K Smyth - 2014*

## DGE format
The DGE output format of DGE functions is a data.frame containing the following columns:\
`gene_id` = Unique NCBI Gene ID\
`DE_log2_FC` = Differential expression expressed as the log2 fold change\
`std.error` = `DE_log2_FC / t.value`\
`t.value` = moderated t-statistic\
`p.value` = p-value\
`adj.p.value` = adjusted p-values calculated using the Benjamini-Hochberg (BH) method.\
\
**Note: DGEs calculated using `lmer` from the `lme4` package do not include the `p.value` and `adj.p.value` columns due to limitations of `lmer`.**

**Example**
```
# gene_id    DE_log2_FC    std.error    t.value     p.value     adj.p.value
# -------------------------------------------------------------------------
# 140        -0.0510758    0.11917072   -0.428593   0.668218    0.924787
# 780        -0.0078762    0.12499224   -0.0630141  0.949755    0.993189
# 2101       -0.0058122    0.08224453   -0.0706701  0.943660    0.986571
# 2597        0.00710451   0.10404212    0.06828492 0.945558    0.999256
# 54812      -0.1922592    0.11767320   -1.8237562  0.102292    0.430841
```

## Gene expression signature
A table containing a subset of the disease DGE corresponding to the n most significant genes.
A typical value of n ranges from 50 to 300.
Only two columns from the original DGE are retained: gene_id and t.value.
The table is structured as a data.frame with two columns:

1. The first column, named "gene_id", contains the gene identifiers.
2. The second column, named "estimate", contains the t-value.

The "estimate" column represents a gene-level statistic derived from the DGE analysis.

This package adopts the convention defined by Koudijs et al. [1], in which the t-value is used as the expression estimate.

In contrast, the implementation by Chen et al. [2] uses the log fold change.\cr

**Note**: For connectivity score computation, the "estimate" value is needed.
The sign of the "estimate" indicates the status of the gene: positive = upregulated, negative = downregulated.

**Example**
```
# gene_id    estimate
# ---------------------
# 140        3.0510758
# 780        -1.0078762
# 2101       -0.1058122
# 2597        0.0071045
# 54812      -2.1922592
```

## Integration with LINCS [4]
The Common Fund’s Library of Integrated Network-based Cellular Signatures (LINCS).

The LINCS data are downloaded from [https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742).

This package provides functions to analyze level 3 data. Level 3 contains RNA analysis
of cancer tissue treated with drugs and control vehicles (DMSO).

Due to the structure of the LINCS database, differential gene expression analysis is performed
for multiple drugs against a single gene.

The LINCS classes use a Linear Mixed Model with the formula: `gene_expression ~ pert_iname + (1 | cell_id) + (1 | rna_plate)` (the fixed effect pert_iname corresponds to the drug name) to compute the DGE.

With this configuration, the DGE is computed for each input drug.

Level 3 data used are:

**extdata/LINCS_gene_info.Rds** (extracted from GSE92742_Broad_LINCS_gene_info.txt)
Gene information.

1. gene_id: gene id
2. gene_symbol: gene symbol
3. lincs_gene_symbol: gene symbol in LINCS database
4. gene_full_name: gene full name
5. is_landmark_gene: indicates whether the gene is a LINCS landmark gene
6. is_best_inferred_gene: indicates whether the gene is well inferred

See example below.

**GSE92742_Broad_LINCS_inst_info.Rds** (extracted from GSE92742_Broad_LINCS_inst_info.txt)

RNA analysis metadata.

1. inst_id: sample id
2. rna_plate: rna plate
3. rna_well: rna well
4. pert_id: id of the perturbagen
5. pert_iname: standardized name of the perturbagen (e.g., drug name)
6. pert_type: category of the perturbation, such as "trt_cp" (chemical treatment) or "ctl_vehicle" (control samples, typically DMSO).
7. pert_dose: concentration of the treatment
8. pert_time: duration of the cellular exposure to the perturbagen (e.g., 6h, 24h, 48h)
9. cell_id: name of the specific cell line used in the assay (e.g., MCF7, PC3, VCAP)

See example below.

**GSE92742_Broad_LINCS_Level3_INF_mlr12k_n1319138x12328.gctx** (not provided with package)

RNA expression data.

For each row in the table GSE92742_Broad_LINCS_inst_info.txt, there is a corresponding column in
GSE92742_Broad_LINCS_Level3_INF_mlr12k_n1319138x12328.gctx. The column name matches the
inst_id column in GSE92742_Broad_LINCS_inst_info.txt.
The table contains only one row, which represents the gene expression level.
Data follow the "normal distribution".

See example below.

**Examples**
```
LINCS_gene_info
# gene_id   gene_symbol   lincs_gene_symbol   gene_full_name    is_landmark_gene
# ------------------------------------------------------------------------------
# 7849      PAX8          PAX8                paired box 8      1
# 2049      EPHB3         EPHB3               EPH receptor B3   0
# 11335     CBX3          CBX3                chromobox 3       0

# is_best_inferred_gene
# ---------------------
# 1
# 1
# 1

GSE92742_Broad_LINCS_inst_info
# inst_id                                 rna_plate            rna_well   pert_id
# -------------------------------------------------------------------------------------
# ASG001_MCF7_24H_X1_B7_DUO52HI53LO:P21   ASG001_MCF7_24H_X1   P21        DMSO
# ASG001_MCF7_24H_X1_B7_DUO52HI53LO:P22   ASG001_MCF7_24H_X1   P22        DMSO
# ASG001_MCF7_24H_X1_B7_DUO52HI53LO:A06   ASG001_MCF7_24H_X1   A06        BRD-A60070924
# ASG001_MCF7_24H_X1_B7_DUO52HI53LO:E06   ASG001_MCF7_24H_X1   E06        BRD-A60070924

# pert_iname        pert_type     pert_dose   pert_time   cell_id
# ---------------------------------------------------------------
# DMSO              ctl_vehicle   0.1         24          MCF7
# DMSO              ctl_vehicle   0.1         24          MCF7
# alpha-estradiol   trt_cp        10          24          MCF7
# alpha-estradiol   trt_cp        10          24          MCF7


GSE92742_Broad_LINCS_Level3_INF_mlr12k_n1319138x12328.gctx
# ASG001_MCF7_24H_X1_B7_DUO52HI53LO:P21   ASG001_MCF7_24H_X1_B7_DUO52HI53LO:P22
# -----------------------------------------------------------------------------
# 9.178                                   9.4127

# ASG001_MCF7_24H_X1_B7_DUO52HI53LO:A06   ASG001_MCF7_24H_X1_B7_DUO52HI53LO:E06
# -----------------------------------------------------------------------------
# 9.2043                                  9.3412

```

## Parallel Processing
The parallel processing is controlled by the `processorCores` instance of the `ProcessorCores` class.
Through  `processorCores` instance, it is possible to retrieve the number of processor cores available on the system and to configure how many cores are used for parallel processing.\
For more information, see the `ProcessorCores` documentation.

## Logging system
Package logging is controlled by the `dgrpLogger` instance of the `DGRPLogger` class.
The `dgrpLogger` instance allows switching the logging mode between `CONSOLE`, `FILE`, or both `CONSOLE` and `FILE`.
It can be used to log activity and data.\
For more information, see the `DGRPLogger` documentation.

## Package overview
The package is entirely developed following the OOP paradigm through the use of the R6 library.

DrugRepurposing provides more than 80 classes that can be composed at will to perform differential gene expression analysis of RNA data and calculate the connectivity score between diseases and drugs in order to find new uses for existing drugs.

The library provides ready-made classes for calculating the connectivity score or DGE, but thanks to the use of the “strategy design pattern”, the behavior of the available classes can be customized.

In the package code, there are more than 100 tests that can be taken as examples for the use of the library itself:
  [https://github.com/carmelo-catalano/DrugRepurposing/tree/devel/tests/testthat](https://github.com/carmelo-catalano/DrugRepurposing/tree/devel/tests/testthat)


## Abstract Classes
In order to make the software structure clearer and facilitate the use of the “strategy design pattern”, pseudo-abstract classes have been created to mimic the behavior of abstract classes found in more structured languages than R, such as C++ or Java.

These classes allow constructor parameters to be “typed” preventing the use of collaborators of the wrong type. As an example, one can see the `lmmVoomDGE` parameter of the constructor of the `GEORNASeqLMMVoomDGE` class.

## Suggestion
  For better use of this package, it is recommended to consult the code at: [https://github.com/carmelo-catalano/DrugRepurposing](https://github.com/carmelo-catalano/DrugRepurposing)

  and to use the tests at:

  [https://github.com/carmelo-catalano/DrugRepurposing/tree/devel/tests/testthat](https://github.com/carmelo-catalano/DrugRepurposing/tree/devel/tests/testthat)

  as guidelines for developing software based on DrugRepurposing.


## Basic Libraries
Statistical analyses are performed using the following libraries: limma, lme4, dream, and the Julia programming language.

## Julia
  To achieve a significant performance boost in the calculation of mixed linear models, the Julia programming language has been utilized. DrugRepurposing is able to invoke Julia's fit function, which makes it possible to compute linear mixed models up to 70 times faster than `lmer`.

  The porting between R and Julia is accomplished through the `JuliaCall` library.
}

## Julia installation
  For Julia environment installation, see: https://julialang.org/.

  After installing Julia on your system, you need to install the Julia packages MixedModels and DataFrames:

   ```
   import Pkg;
   Pkg.add("MixedModels")
   Pkg.add("DataFrames")
   ```

## Troubleshooting Julia Integration with R
  If your R environment is unable to locate the Julia installation, you may need to configure the JULIA_HOME environment variable. Try one or both of the following instructions:

1.   Option 1: Set JULIA_HOME within R.\
  options(JULIA_HOME = "/your-julia-installation-folder/bin/").\
  This command sets the JULIA_HOME option within your current R session, pointing to the Julia executable. To make it persistent, you can add this line to your .Rprofile file.

2. Option 2: Set JULIA_HOME in your system environment.\
    Bash\
    export JULIA_HOME=/your-julia-installation-folder/bin/.\
    This command sets the JULIA_HOME environment variable for your current shell session. You might need to add this line to your shell's configuration file (e.g., .bashrc, .zshrc) to make it permanent.


## Main classes
**Drug repurposing**
```
DichotomicDrugRepurpose
DichotomicVoomDrugRepurpose
GEORNASeqDichotomicVoomDrugRepurpose
LMMDrugRepurpose
LMMVoomDrugRepurpose
GEORNASeqLMMVoomDrugRepurpose
```

**Connectivity score**
```
BinChenDiseaseSignatureDrugListConnectivityScoreParallelFacade
BinChenDiseaseSignatureDrugListConnectivityScoreSyncFacade
```

**Dichotomic differential gene expression**
```
DichotomicDGE
DichotomicVoomDGE
GEORNASeqDichotomicVoomDGE
```

**LMM differential gene expression**
```
LMMDreamDGE
LMMJuliaDGE
LMMLmerDGE
LMMVoomDreamDGE
LMMVoomJuliaDGE
LMMVoomLmerDGE
GEORNASeqLMMVoomDreamDGE
GEORNASeqLMMVoomJuliaDGE
GEORNASeqLMMVoomLmerDGE
LINCSRDSLMMDrugDGEJulia
LINCSRDSLMMDrugDGEJuliaClusters
LINCSRDSLMMDrugDGELmer
```

**Metanalysis**
```
DGEMetanalysisByMatrix
```

**LINCS [3] GCTX archive export**
```
LINCSExport
```

## References
1.	Koudijs K.K.M. et al. Validation of transcriptome signature reversion for drug repurposing in oncology. Briefings in Bioinformatics, 2022.
2.	Chen B. et al. Reversal of cancer gene expression correlates with drug efficacy and reveals therapeutic targets. Nature Communications, 2017; 8:16022.
3.	Lamb J. et al. The Connectivity Map: using gene-expression signatures to connect small molecules, genes, and disease. Science, 2006; 313(5795):1929–1935.
4.	Keenan A.B. et al. The Library of Integrated Network-Based Cellular Signatures NIH Program: System-Level Cataloging of Human Cells Response to Perturbations. Cell Systems, 2018; 6.


#
# Examples

  Here you can find two complete examples built on real data from NCBI ([https://www.ncbi.nlm.nih.gov/](https://www.ncbi.nlm.nih.gov/)):

[https://github.com/carmelo-catalano/DrugRepurposing/tree/devel/examples](https://github.com/carmelo-catalano/DrugRepurposing/tree/devel/examples)

** **

  Below you will find a complete example built on invented data.


  DrugRepurposing package version: `r packageVersion("DrugRepurposing")`

```{r setup, include=FALSE}
library(DrugRepurposing)
options(warn = -1)
```
```{r}
library(DrugRepurposing)
rna_seq_drug_1 <- matrix(
  c(
    262, 281897, 463, 12301, 284, 2966, 4842, 9966,
    116, 337026, 261, 3277, 132, 2006, 3699, 12406,
    111, 111736, 193, 862, 88, 1371, 2536, 1531,
    162, 219471, 281, 5990, 280, 2241, 4527, 12941
  ),
  nrow = 8,
  ncol = 4,
  dimnames = list(
    c(1, 2, 3, 9, 10, 11, 12, 13),
    c("GSM1", "GSM2", "GSM3", "GSM4")
  )
)

sample_01_map_drug_1 <- "1X01"
# How the map is applied:
#           1      X      0      1
# GeneID   GSM1   GSM2   GSM3   GSM4
# 1        262    281897 463    12301
# ...
#
#
# GSM1, GSM4 = IPF; GSM3 = control; GSM2 = excluded

filter_by_protein_coding <- FALSE

dichotomicVoomDGE <- DichotomicVoomDGE$new()
drug_1_DGE <- dichotomicVoomDGE$compute(rna_seq_drug_1, sample_01_map_drug_1, filter_by_protein_coding)

drug_1 <- drug_1_DGE[, c("gene_id", "t.value")]
colnames(drug_1)[2] <- "estimate"

rna_seq_drug_2 <- matrix(
  c(
    146, 247, 504, 938, 627, 733, 844,
    245, 609, 108, 608, 928, 534, 48,
    809, 765, 979, 832, 915, 981, 395,
    644, 124, 830, 296, 738, 120, 235,
    26, 60, 927, 622, 694, 940, 859,
    712, 540, 448, 822, 195, 100, 814
  ),
  nrow = 6,
  ncol = 7,
  dimnames = list(
    c(1, 2, 3, 9, 10, 11),
    c("GSM1", "GSM2", "GSM3", "GSM4", "GSM5", "GSM6", "GSM7")
  )
)

sample_01_map_drug_2 <- "X110011"
# How the map is applied:
#           X      1      1      0      0      1      1
# GeneID   GSM1   GSM2   GSM3   GSM4   GSM5   GSM6   GSM7
# 1        146    247    504    938    627    733    844
# ...
#

drug_2_DGE <- dichotomicVoomDGE$compute(rna_seq_drug_2, sample_01_map_drug_2, filter_by_protein_coding)
drug_2 <- drug_2_DGE[, c("gene_id", "t.value")]
colnames(drug_2)[2] <- "estimate"

drugs_rna_data <- list()
drugs_rna_data[["drug_1"]] <- drug_1
drugs_rna_data[["drug_2"]] <- drug_2

drug_signatures <- data.frame(
  name = c("drug_1", "drug_2"),
  signature_reference = c("drug_1", "drug_2")
)

EasyDrugSignatureLoader <- R6Class(
  "EasyDrugSignatureLoader",
  inherit = DrugSignatureLoaderAbstract,
  public = list(
    initialize = function(drugs_rna_data) {
      private$drugs_rna_data <- drugs_rna_data
    },
    load = function(drug_name) {
      return(private$drugs_rna_data[[drug_name]])
    }
  ),
  private = list(
    drugs_rna_data = NA
  )
)


meanThresholdGeneFilter <- MeanThresholdGeneFilter$new(1)
easyDrugSignatureLoader <- EasyDrugSignatureLoader$new(drugs_rna_data)

disease_rna_seq <- matrix(
  c(
    96, 311, 142, 428, 781,
    466, 973, 193, 621, 405,
    537, 179, 65, 486, 267,
    308, 721, 714, 761, 605,
    140, 77, 727, 116, 8,
    460, 520, 645, 881, 328,
    866, 887, 730, 157, 26,
    825, 588, 113, 851, 667
  ),
  nrow = 8,
  ncol = 5,
  dimnames = list(
    c(1, 2, 3, 9, 10, 11, 12, 13),
    c("GSM1", "GSM2", "GSM3", "GSM4", "GSM5")
  )
)

disease_sample_01_map <- "1X011"
drugs_genes <- c(1, 2, 3, 9, 10, 11)


dichotomicVoomDrugRepurpose <- DichotomicVoomDrugRepurpose$new(
  meanThresholdGeneFilter,
  easyDrugSignatureLoader
)
dichotomicVoomDrugRepurpose$compute(
  rna_seq = disease_rna_seq,
  sample_01_map = disease_sample_01_map,
  drug_signatures = drug_signatures,
  drugs_genes = drugs_genes,
  random_distribution_size = 10, # 10 is a poor value used only to speed up
                                 # computation; an appropriate value is 10^5
  disease_name = "IPF",
  filter_by_protein_coding = FALSE,
  parallel_computation = FALSE,
  signature_mapper_parameter = 6 # This parameter typically represents the most significant genes of the disease,
                                 # and drug repurposing is computed only on this subset.
)

```