---
title: "GEO RNA Seq LMM Voom DGE by Lmer"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
 %\VignetteIndexEntry{GEO RNA Seq LMM Voom DGE by Lmer}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
# Description
Derived class of `GEORNASeqLMMVoomDGEAbstract`.
Computes the Voom Linear Mixed Model (LMM) differential gene expression (DGE) analysis using the `lmer` function from the `lme4` package.

The input consists of two groups of RNA-Seq data from tissues under different conditions: control/altered state. For example, a disease-affected group and a control group.

Unlike the dichotomous case, where groups are divided only into "control/altered state," additional conditions exist under which the experiments were conducted. For instance, multiple RNA plates were used, or tissues originated from different parts of the body.

RNA data must follow a negative binomial distribution, typically data from RNA-Seq-like analyses.

The input RNA-Seq data file is sourced from NCBI [https://www.ncbi.nlm.nih.gov/](https://www.ncbi.nlm.nih.gov/).

```{r setup, include=FALSE}
library(DrugRepurposing)
options(warn = -1)
```
```{r}
library(DrugRepurposing)

tsvLines <- list(
  "GeneID	GSM1	GSM2	GSM3	GSM4	GSM5	GSM6	GSM7	GSM8	GSM9	GSM10	GSM11	GSM12",
  "203228	632	944	219	547	895	610	369	427	763	3	772	980",
  "6647	51	659	275	911	449	509	742	410	935	66	238	358",
  "2521	323	681	521	947	133	492	459	224	876	692	927	836"
)
tsv <- tsvLines[1]
for (i in 2:4)
  tsv <- paste(tsv, tsvLines[i], sep = "\n")
rna_seq_data_filename <- tempfile()
writeLines(tsv, con = rna_seq_data_filename)

sample_id <- c("GSM1", "GSM2", "GSM3", "GSM4", "GSM5", "GSM6",
               "GSM7", "GSM8", "GSM9", "GSM10", "GSM11", "GSM12")
tissue_status <- c("ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS",
                   "ALS", "ALS", "Control", "Control")
# The first level must be "Control" state
tissue_status_factor <- factor(tissue_status, levels = c("Control", "ALS"))
rna_seq_metadata <- data.frame(
  sample_id = sample_id,
  tissue = c("Spinal Cord Lumbar", "Spinal Cord Lumbar", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Cerebellum", "Cerebellum", "Cerebellum",
             "Cerebellum", "Cerebellum", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Cerebellum"),
  tissue_status = tissue_status_factor
)

random_effect_column_names <- "tissue"
sample_status_column_name <- "tissue_status"
formula <- ~tissue_status + (1 | tissue)

# In this example, we have limited data, so the random-effects filter is disabled.
# In a typical DGE analysis, disabling this filter is not recommended.
lmmGeneFilter <- LMMLowCountsGeneFilter$new(random_effects_level_threshold = 0)
geoRNASeqLMMVoomDGELmer <- GEORNASeqLMMVoomDGELmer$new(lmmGeneFilter)
geoRNASeqLMMVoomDGELmer$compute(
  rna_seq_data_filename = rna_seq_data_filename,
  rna_seq_metadata = rna_seq_metadata,
  formula =formula,
  random_effect_column_names = random_effect_column_names,
  counts_filter_column_name = sample_status_column_name,
  filter_by_protein_coding = FALSE
)

unlink(rna_seq_data_filename)
```
