---
title: "GEO RNA Seq LMM Voom DGE by Dream With eBayes"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
 %\VignetteIndexEntry{GEO RNA Seq LMM Voom DGE by Dream With eBayes}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
# Description
Derived class of `GEORNASeqLMMVoomDGEAbstract`.

Computes the Voom Linear Mixed Model (LMM) differential gene expression (DGE) analysis using the `dream` function from the `variancePartition` package.

After the `dream` computation is completed "Empirical Bayes Statistics for Differential Expression" are applied (`variancePartition::eBayes`).

The input consists of two groups of RNA-Seq data from tissues under different conditions: control/altered state. For example, a disease-affected group and a control group.

Unlike the dichotomous case, where groups are divided only into "control/altered state," additional conditions exist under which the experiments were conducted. For instance, multiple RNA plates were used, or tissues originated from different parts of the body.

RNA data must follow a negative binomial distribution, typically data from RNA-Seq-like analyses.

The input RNA-Seq data file is sourced from NCBI [https://www.ncbi.nlm.nih.gov/](https://www.ncbi.nlm.nih.gov/).

```{r setup, include=FALSE}
library(DrugRepurposing)
options(warn = -1)
```
```{r}
library(DrugRepurposing)

tsvLines <- list(
  "GeneID	GSM2433098	GSM2433099	GSM2433100	GSM2433101	GSM2433102	GSM2433103	GSM2433104	GSM2433109",
  "1	817	480	513	497	1055	409	661	788",
  "2	451	315	489	620	762	244	572	234",
  "9	792	608	560	585	952	436	823	981",
  "12	220	244	283	404	434	196	314	225",
  "13	142	229	240	350	332	163	239	324",
  "14	682	501	354	669	721	520	550	755",
  "21	230	133	83	59	127	91	86	100",
  "25	1022	605	246	368	611	588	654	811"
)
tsv <- ""
for (i in 1:9)
  tsv <- paste(tsv, tsvLines[i], sep = "\n")
rna_seq_data_filename <- tempfile()
writeLines(tsv, con = rna_seq_data_filename)

tissue_status <- c("ALS Spectrum MND", "ALS Spectrum MND", "ALS Spectrum MND",
                    "Non-Neurological Control", "ALS Spectrum MND", "ALS Spectrum MND",
                    "Non-Neurological Control")
# The first level must be "Control" state
tissue_status_factor <- factor(tissue_status, levels = c("Non-Neurological Control", "ALS Spectrum MND"))

rna_seq_metadata <- data.frame(
  sample_id = c("GSM2433098", "GSM2433099", "GSM2433100", "GSM2433101",
                "GSM2433102", "GSM2433103", "GSM2433104"),
  tissue = c("Spinal Cord Lumbar", "Spinal Cord Lumbar", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Spinal Cord Lumbar", "Cerebellum", "Cerebellum"),
  subject_id = c("NEUEL133AK6", "NEUYV496XLP", "NEUZU200WEQ", "NEUEC006FND",
                 "NEUKM699KKH", "NEUJZ255ZC5", "NEUJZ233ZL3"),
  tissue_status = tissue_status_factor
)
# rna_seq_metadata structure:
#
# sample_id       tissue                subject_id     tissue_status
# GSM2433098      Spinal Cord Lumbar    NEUEL133AK6   ALS Spectrum MND
# GSM2433099      Spinal Cord Lumbar    NEUYV496XLP   ALS Spectrum MND
# GSM2433100      Spinal Cord Lumbar    NEUZU200WEQ   ALS Spectrum MND
# GSM2433101      Spinal Cord Lumbar    NEUEC006FND   Non-Neurological Control
# GSM2433102      Spinal Cord Lumbar    NEUKM699KKH   ALS Spectrum MND
# GSM2433103      Cerebellum            NEUJZ255ZC5   ALS Spectrum MND
# GSM2433104      Cerebellum            NEUJZ233ZL3   Non-Neurological Control
# sample_id column is used as a filter for rna_seq_data:
# rna_seq_data <- rna_seq_data[, rna_seq_metadata$sample_id]
# In this example, the GSM2433109 column of rna_seq_data will be discarded before computation.

random_effect_column_names <- "tissue"
sample_status_column_name <- "tissue_status"
formula <- ~tissue_status + (1 | tissue)

# In this example, we have limited data, so the random-effects filter is disabled.
# In a typical DGE analysis, disabling this filter is not recommended.
lmmGeneFilter <- LMMLowCountsGeneFilter$new(random_effects_level_threshold = 0)
geoRNASeqLMMVoomDGEDreamWithEBayes <- GEORNASeqLMMVoomDGEDreamWithEBayes$new(lmmGeneFilter)
geoRNASeqLMMVoomDGEDreamWithEBayes$compute(
  rna_seq_data_filename = rna_seq_data_filename,
  rna_seq_metadata = rna_seq_metadata,
  formula = formula,
  random_effect_column_names = random_effect_column_names,
  counts_filter_column_name = sample_status_column_name,
  filter_by_protein_coding = FALSE
)

unlink(rna_seq_data_filename)
```
