---
title: "GEO RNA Seq Dichotomic Voom Differential Gene Expression analysis (DGE)"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
  %\VignetteIndexEntry{GEO RNA Seq Dichotomic Voom Differential Gene Expression analysis (DGE)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Description
Computes the differential gene expression analysis.

The input consists of two groups of RNA-Seq data from tissues under different conditions, for
 example, a disease-affected group and a control group. RNA data must follow a negative
 binomial distribution, typically data from RNA-Seq-like analyses.

For RNA data following a normal distribution, such as microarray data, see `DichotomicDGE`.

The input RNA-Seq data file is sourced from NCBI
[https://www.ncbi.nlm.nih.gov/](https://www.ncbi.nlm.nih.gov/).

```{r setup, include=FALSE}
library(DrugRepurposing)
```
```{r}
library(DrugRepurposing)
rna_seq_filename <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts&acc=GSE92592&file=GSE92592_raw_counts_GRCh38.p13_NCBI.tsv.gz"
# File content:
# GeneID   GSM1    GSM2    GSM3    GSM4
# 1        262     116     111     162
# 2        281897  337026  111736  219471
# 3        463     261     193     281
# 9        12301   3277    862     5990
# 10       284     132     88      280
# 11       2966    2006    1371    2241
# 12       4842    3699    2536    4527
# 13       9966    12406   1531    12941

sample_01_map <- "0X10"

sample_01_map <- "0X10"
# How the map is applied:
#           0      X      1      0
# GeneID   GSM1   GSM2   GSM3   GSM4
# 1        262    116    111    162
# ...
#
#
# GSM1, GSM4 = IPF; GSM3 = control; GSM2 = excluded

filter_by_protein_coding <- FALSE

geoRNASeqDichotomicVoomDGE <- GEORNASeqDichotomicVoomDGE$new()

geoRNASeqDichotomicVoomDGE$compute(
  rna_seq_filename, sample_01_map, filter_by_protein_coding
)
```
