---
title: "GEO RNA Seq Dichotomic Voom Drug Repurpose"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
 %\VignetteIndexEntry{GEO RNA Seq Dichotomic Voom Drug Repurpose}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
# Description
Computes the connectivity score between an RNA-Seq dataset of a disease and a list of drugs.

This class computes the differential gene expression (DGE) of the disease and then
computes the connectivity scores. The disease DGE is computed by `GEORNASeqDichotomicVoomDGE`.
The connectivity scores are computed by `BinChenDiseaseDGEDrugListConnectivityScore`.

RNA data must follow a negative binomial distribution, typically data from RNA-Seq-like analyses.

The input RNA-Seq data file is sourced from NCBI [https://www.ncbi.nlm.nih.gov/](https://www.ncbi.nlm.nih.gov/)..

```{r setup, include=FALSE}
library(DrugRepurposing)
```
```{r}
library(DrugRepurposing)
rna_seq_data_filename <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts&acc=GSE92592&file=GSE92592_raw_counts_GRCh38.p13_NCBI.tsv.gz"
# File content example:
# GeneID   GSM1    GSM2    GSM3    GSM4
# 1        262     116     111     162
# 2        281897  337026  111736  219471
# 3        463     261     193     281
# 9        12301   3277    862     5990
# 10       284     132     88      280
# 11       2966    2006    1371    2241
# 12       4842    3699    2536    4527
# 13       9966    12406   1531    12941

sample_01_map <- "1X01"

# How the map is applied:
#           1      X      0      1
# GeneID   GSM1   GSM2   GSM3   GSM4
# 1        262    116    111    162
# ...
#
#
# GSM1, GSM4 = IPF; GSM3 = control; GSM2 = excluded

drugs_genes <- c(1, 2, 9, 12, 13, 14, 21, 25)

drug_signatures <- data.frame(
  name = c("A-23187", "A-443644", "AG-490", "AG-494"),
  signature_reference = c("A-23187", "A-443644", "AG-490", "AG-494")
)

A_23187_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 31),
  estimate = c(1.98564, -1.51579, 0.33099, -0.935718, -3.818529,
               1.848352, -4.332752, 3.591474, -0.345235423)
)

A_443644_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 6158),
  estimate = c(-1.55774, -4.4155, 1.21015, 4.8705936, -0.1259465,
               -2.6003531, -0.3530122, -3.5426363, -3.252352)
)

AG_490_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 37),
  estimate = c(-1.35143, 0.58724, -1.31592, 0.22267776, -0.45726445,
               -2.80993289, -0.05550198, 1.76931022, 0.1423542)
)

AG_494_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 6158),
  estimate = c(-3.43357, -0.14896, 0.90733, 3.084164, -2.594041,
               4.304979, -4.944015, -2.075178, 1.4523542)
)

drugs_rna_data <- list()
drugs_rna_data[["A-23187"]] <- A_23187_drug
drugs_rna_data[["A-443644"]] <- A_443644_drug
drugs_rna_data[["AG-490"]] <- AG_490_drug
drugs_rna_data[["AG-494"]] <- AG_494_drug


EasyDrugSignatureLoader <- R6Class(
  "EasyDrugSignatureLoader",
  inherit = DrugSignatureLoaderAbstract,
  public = list(
    initialize = function(drugs_rna_data) {
      private$drugs_rna_data <- drugs_rna_data
    },
    load = function(drug_name) {
      return(private$drugs_rna_data[[drug_name]])
    }
  ),
  private = list(
    drugs_rna_data = NA
  )
)

easyDrugSignatureLoader <- EasyDrugSignatureLoader$new(drugs_rna_data)

geoRNASeqDichotomicVoomDrugRepurpose <- GEORNASeqDichotomicVoomDrugRepurpose$new(
  drugSignatureLoader = easyDrugSignatureLoader)
geoRNASeqDichotomicVoomDrugRepurpose$compute(
  rna_seq_data_filename = rna_seq_data_filename,
  sample_01_map = sample_01_map,
  drug_signatures = drug_signatures,
  drugs_genes = drugs_genes,
  random_distribution_size = 10, # 10 is a poor value used only to speed up
                                 # computation; an appropriate value is 10^5
  disease_name = "IPF",
  parallel_computation = FALSE,
  filter_by_protein_coding = FALSE,
  signature_mapper_parameter = 8
)

```
