---
title: "GEO RNA Seq LMM Voom Drug Repurpose"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
 %\VignetteIndexEntry{GEO RNA Seq LMM Voom Drug Repurpose}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
# Description
Given a disease RNA-Seq analysis and a list of drug DGEs, `GEORNASeqLMMVoomDrugRepurpose`
computes the connectivity score between the disease and the drug DGEs.
\cr
`GEORNASeqLMMVoomDrugRepurpose` calculates differential gene expression from RNA-seq data using a
Linear Mixed Model, and then computes the connectivity score between the disease DGE and the drug DGEs.
\cr
This class composes the class BinChenDiseaseDGEDrugListConnectivityScore and an implementation
of `GEORNASeqLMMVoomDGEAbstract`.

RNA data must follow a negative binomial distribution, typically data from RNA-Seq-like analyses.

The input RNA-Seq data file is sourced from NCBI [https://www.ncbi.nlm.nih.gov/](https://www.ncbi.nlm.nih.gov/).

```{r setup, include=FALSE}
library(DrugRepurposing)
options(warn = -1)
```
```{r}
library(DrugRepurposing)

tsvLines <- list(
  "GeneID	GSM2433098	GSM2433099	GSM2433100	GSM2433101	GSM2433102	GSM2433103	GSM2433104	GSM2433109",
  "1	817	480	513	497	1055	409	661	788",
  "2	451	315	489	620	762	244	572	234",
  "9	792	608	560	585	952	436	823	981",
  "12	220	244	283	404	434	196	314	225",
  "13	142	229	240	350	332	163	239	324",
  "14	682	501	354	669	721	520	550	755",
  "21	230	133	83	59	127	91	86	100",
  "25	1022	605	246	368	611	588	654	811"
)
tsv <- ""
for (i in 1:9)
  tsv <- paste(tsv, tsvLines[i], sep = "\n")
rna_seq_data_filename <- tempfile()
writeLines(tsv, con = rna_seq_data_filename)

rna_seq_metadata <- data.frame(
  sample_id = c("GSM2433098", "GSM2433099", "GSM2433100", "GSM2433101",
                "GSM2433102", "GSM2433103", "GSM2433104"),
  tissue = c("Spinal Cord Lumbar", "Spinal Cord Lumbar", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Spinal Cord Lumbar", "Cerebellum", "Cerebellum"),
  subject_id = c("NEUEL133AK6", "NEUYV496XLP", "NEUZU200WEQ", "NEUEC006FND",
                 "NEUKM699KKH", "NEUJZ255ZC5", "NEUJZ233ZL3"),
  tissue_status = c("ALS Spectrum MND", "ALS Spectrum MND", "ALS Spectrum MND",
                    "Non-Neurological Control", "ALS Spectrum MND", "ALS Spectrum MND",
                    "Non-Neurological Control")
)
# rna_seq_metadata structure:
#
# sample_id       tissue                subject_id     tissue_status
# GSM2433098      Spinal Cord Lumbar    NEUEL133AK6   ALS Spectrum MND
# GSM2433099      Spinal Cord Lumbar    NEUYV496XLP   ALS Spectrum MND
# GSM2433100      Spinal Cord Lumbar    NEUZU200WEQ   ALS Spectrum MND
# GSM2433101      Spinal Cord Lumbar    NEUEC006FND   Non-Neurological Control
# GSM2433102      Spinal Cord Lumbar    NEUKM699KKH   ALS Spectrum MND
# GSM2433103      Cerebellum            NEUJZ255ZC5   ALS Spectrum MND
# GSM2433104      Cerebellum            NEUJZ233ZL3   Non-Neurological Control
# sample_id column is used as a filter for rna_seq_data:
# rna_seq_data <- rna_seq_data[, rna_seq_metadata$sample_id]
# In this example, the GSM2433109 column of rna_seq_data will be discarded before computation.

drugs_genes <- c(1, 2, 9, 12, 13, 14, 21, 25)

drugs <- data.frame(
  name = c("A-23187", "A-443644", "AG-490", "AG-494"),
  filename = c("A-23187", "A-443644", "AG-490", "AG-494")
)

A_23187_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 31),
  estimate = c(1.98564, -1.51579, 0.33099, -0.935718, -3.818529,
               1.848352, -4.332752, 3.591474, -0.345235423)
)

A_443644_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 6158),
  estimate = c(-1.55774, -4.4155, 1.21015, 4.8705936, -0.1259465,
               -2.6003531, -0.3530122, -3.5426363, -3.252352)
)

AG_490_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 37),
  estimate = c(-1.35143, 0.58724, -1.31592, 0.22267776, -0.45726445,
               -2.80993289, -0.05550198, 1.76931022, 0.1423542)
)

AG_494_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 6158),
  estimate = c(-3.43357, -0.14896, 0.90733, 3.084164, -2.594041,
               4.304979, -4.944015, -2.075178, 1.4523542)
)

drugs_rna_data <- list()
drugs_rna_data[["A-23187"]] <- A_23187_drug
drugs_rna_data[["A-443644"]] <- A_443644_drug
drugs_rna_data[["AG-490"]] <- AG_490_drug
drugs_rna_data[["AG-494"]] <- AG_494_drug


EasyDrugSignatureLoader <- R6Class(
  "EasyDrugSignatureLoader",
  inherit = DrugSignatureLoaderAbstract,
  public = list(
    initialize = function(drugs_rna_data) {
      private$drugs_rna_data <- drugs_rna_data
    },
    load = function(drug_name) {
      return(private$drugs_rna_data[[drug_name]])
    }
  ),
  private = list(
    drugs_rna_data = NA
  )
)

easyDrugSignatureLoader <- EasyDrugSignatureLoader$new(drugs_rna_data)

geoRNASeqLMMVoomDGELmer <- GEORNASeqLMMVoomDGEDream$new()
geoRNASeqLMMVoomDrugRepurpose <- GEORNASeqLMMVoomDrugRepurpose$new(
  geoRNASeqLMMVoomDGE = geoRNASeqLMMVoomDGELmer,
  drugSignatureLoader = easyDrugSignatureLoader
)

geoRNASeqLMMVoomDrugRepurpose$compute(
  rna_seq_data_filename = rna_seq_data_filename,
  rna_seq_metadata = rna_seq_metadata,
  formula = ~tissue_status + (1 | tissue),
  drugs = drugs,
  drugs_genes = drugs_genes,
  random_distribution_size = 10,
  disease_name = "ALS",
  parallel_computation = FALSE,
  filter_by_protein_coding = FALSE,
  signature_mapper_parameter = 8
)

unlink(rna_seq_data_filename)
```
