---
title: "Linear Mixed Model DGE by Lmer"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
 %\VignetteIndexEntry{Linear Mixed Model DGE by Lmer}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
# Description
Derived class of `LMMDGEAbstract`.

Computes differential gene expression analysis using the Linear Mixed Model (LMM).

The LMM computation is performed using the `lmer` function (library `lme4`).

The input consists of two groups of RNA data analysis from tissues under different conditions: control/altered state.
For example, a disease-affected group and a control group.

Unlike the dichotomous case, where groups are divided only into "control/altered state", additional conditions exist under which the experiments were conducted. For instance, multiple RNA plates were used, or tissues originated from different parts of the body.
RNA data must follow "normal distribution", which is typical for RNA microarray analysis.

For RNA data following a "negative binomial distribution", refer to `LMMVoomDGEDream`, `LMMVoomDGEJulia`, or `LMMVoomDGELmer`.



```{r setup, include=FALSE}
library(DrugRepurposing)
options(warn = -1)
```
```{r}
library(DrugRepurposing)
sample_id <- c("GSM1", "GSM2", "GSM3", "GSM4", "GSM5", "GSM6",
               "GSM7", "GSM8", "GSM9", "GSM10", "GSM11", "GSM12")
rna_data <- matrix(
  # rna microarray analysis
  c(2.9972499, 4.2469098, 19.0690947, 7.5957615, -6.5289474, 36.6519995,
    12.2326273, -1.2400751, -12.0185859, 9.6117539, -9.7031797, 11.9692450,
    -17.2317513, 13.3941123, 19.7100277, -9.7931619, 15.3568206, 26.0890177,
    8.9792826, 4.0547337, 36.0595866, -7.2987920, 19.1745696, 30.5492311,
    -10.6683999, 11.9134902, 4.3919955, -18.0423696, 0.4506173, 15.8493678,
    -4.6110365, -7.3977498, -9.0876729, -10.1055128, -3.1825579, -2.4332119),
  nrow = 3,
  ncol = 12,
  dimnames = list(
    # gene IDs
    c(203228, 6647, 2521),
    # sample_id
    sample_id
  )
)
# the first column represents the row names (gene IDs)
# rna_data =
#          GSM1        GSM2        GSM3         GSM4          GSM5         ...
# 203228   2.9972499   4.2469098   19.0690947   7.5957615     -6.5289474   ...
# 6647     9.6117539   -9.7031797  11.9692450   -17.2317513   13.3941123   ...
# 2521     8.9792826   4.0547337   36.0595866   -7.2987920    19.1745696   ...

tissue_status <- c("ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS",
                   "ALS", "ALS", "Control", "Control")
tissue_status_factor <- factor(tissue_status, levels = c("Control", "ALS"))
rna_metadata <- data.frame(
  sample_id = sample_id,
  tissue = c("Spinal Cord Lumbar", "Spinal Cord Lumbar", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Cerebellum", "Cerebellum", "Cerebellum",
             "Cerebellum", "Cerebellum", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Cerebellum"),
  tissue_status = tissue_status_factor
)
formula <- ~tissue_status + (1 | tissue)

lmmDGELmer <- LMMDGELmer$new()
lmmDGELmer$compute(
  rna_data = rna_data,
  rna_metadata = rna_metadata,
  formula = formula,
  filter_by_protein_coding = FALSE
)
```
