---
title: "LMM Drug Repurpose"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
 %\VignetteIndexEntry{LMM Drug Repurpose}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
# Description
Given a disease RNA data analysis and a list of drug DGEs, `LMMDrugRepurpose` computes the connectivity score between the disease and the drugs.

`LMMDrugRepurpose` calculates differential gene expression from RNA data using a
Linear Mixed Model, and then computes the connectivity score between the disease DGE and the drug DGEs.

This class creates an instance of `LMMDrugRepurposeCore` and uses it for the computation.

RNA data must be on the "normal distribution", usually data from microarray like analysis.

For RNA data on negative binomial distribution like Illumina RNA-Seq see  `LMMVoomDrugRepurpose`.
```{r setup, include=FALSE}
library(DrugRepurposing)
```
```{r}
library(DrugRepurposing)

sample_id <- c("GSM1", "GSM2", "GSM3", "GSM4", "GSM5", "GSM6",
               "GSM7", "GSM8", "GSM9", "GSM10", "GSM11", "GSM12")
rna_data <- matrix(
  # rna microarray analysis
  c(2.9972499, 4.2469098, 19.0690947, 7.5957615, -6.5289474, 36.6519995,
    12.2326273, -1.2400751, -12.0185859, 9.6117539, -9.7031797, 11.9692450,
    -17.2317513, 13.3941123, 19.7100277, -9.7931619, 15.3568206, 26.0890177,
    8.9792826, 4.0547337, 36.0595866, -7.2987920, 19.1745696, 30.5492311,
    -10.6683999, 11.9134902, 4.3919955, -18.0423696, 0.4506173, 15.8493678,
    -4.6110365, -7.3977498, -9.0876729, -10.1055128, -3.1825579, -2.4332119),
  nrow = 3,
  ncol = 12,
  dimnames = list(
    # gene IDs
    c(203228, 6647, 2521),
    # sample_id
    sample_id
  )
)
# the first column represents the row names (gene IDs)
# rna_data =
#          GSM1        GSM2        GSM3         GSM4          GSM5         ...
# 203228   2.9972499   4.2469098   19.0690947   7.5957615     -6.5289474   ...
# 6647     9.6117539   -9.7031797  11.9692450   -17.2317513   13.3941123   ...
# 2521     8.9792826   4.0547337   36.0595866   -7.2987920    19.1745696   ...

tissue_status <- c("ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS",
                   "ALS", "ALS", "Control", "Control")
tissue_status_factor <- factor(tissue_status, levels = c("Control", "ALS"))
rna_metadata <- data.frame(
  sample_id = sample_id,
  tissue = c("Spinal Cord Lumbar", "Spinal Cord Lumbar", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Cerebellum", "Cerebellum", "Cerebellum",
             "Cerebellum", "Cerebellum", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Cerebellum"),
  tissue_status = tissue_status_factor
)
# rna_metadata =
# sample_id   tissue                  tissue_status
# GSM1        Spinal Cord Lumbar      ALS
# GSM2        Spinal Cord Lumbar      ALS
# GSM3        Spinal Cord Lumbar      ALS
# GSM4        Spinal Cord Lumbar      ALS
# GSM5        Cerebellum              ALS
# GSM6        Cerebellum              ALS
# GSM7        Cerebellum              ALS
# GCS8        Cerebellum              ALS
# GSM9        Cerebellum              ALS
# GSM10       Spinal Cord Lumbar      ALS
# GSM11       Spinal Cord Lumbar      Control
# GSM12       Cerebellum              Control

formula <- ~tissue_status + (1 | tissue)

drugs <- data.frame(
  name = c("A-23187", "A-443644", "AG-490", "AG-494"),
  filename = c("A-23187", "A-443644", "AG-490", "AG-494")
)

A_23187_drug <- data.frame(
  gene_id = c(203228, 6647, 2521),
  estimate = c(1.98564, -1.51579, 0.33099)
)

A_443644_drug <- data.frame(
  gene_id = c(203228, 6647, 2521),
  estimate = c(-1.55774, -4.4155, 1.21015)
)

AG_490_drug <- data.frame(
  gene_id = c(203228, 6647, 2521),
  estimate = c(-1.35143, 0.58724, -1.31592)
)

AG_494_drug <- data.frame(
  gene_id = c(203228, 6647, 2521),
  estimate = c(-3.43357, -0.14896, 0.90733)
)

drugs_rna_data <- list()
drugs_rna_data[["A-23187"]] <- A_23187_drug
drugs_rna_data[["A-443644"]] <- A_443644_drug
drugs_rna_data[["AG-490"]] <- AG_490_drug
drugs_rna_data[["AG-494"]] <- AG_494_drug


drugs_genes <- c(203228, 6647, 2521)

EasyDrugSignatureLoader <- R6Class(
  "EasyDrugSignatureLoader",
  inherit = DrugSignatureLoaderAbstract,
  public = list(
    initialize = function(drugs_rna_data) {
      private$drugs_rna_data <- drugs_rna_data
    },
    load = function(drug_name) {
      return(private$drugs_rna_data[[drug_name]])
    }
  ),
  private = list(
    drugs_rna_data = NA
  )
)

easyDrugSignatureLoader <- EasyDrugSignatureLoader$new(drugs_rna_data)
lmmDGE <- LMMDGELmer$new() # Alternatives: LMMDGEDream, LMMDGEJulia
lmmDrugRepurpose <- LMMDrugRepurpose$new(lmmDGE, easyDrugSignatureLoader)
lmmDrugRepurpose$compute(
  rna_data = rna_data,
  rna_metadata = rna_metadata,
  formula = formula,
  drugs = drugs,
  drugs_genes = drugs_genes,
  random_distribution_size = 10, # 10 is a poor value used only to speed up
                                 # computation; an appropriate value is 10^5
  disease_name = "ALS",
  parallel_computation = FALSE,
  filter_by_protein_coding = FALSE,
  signature_mapper_parameter = 3
)

```