---
title: "Linear Mixed Model Voom DGE by Lmer"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
 %\VignetteIndexEntry{Linear Mixed Model Voom DGE by Lmer}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
# Description
Derived class of `LMMVoomDGEAbstract`.

Computes the Voom Linear Mixed Model (LMM) differential gene expression (DGE) analysis using the `lmer` function from the `lme4` package.

It relies on three collaborators: `LMMVoom`, `DGEMapper`, and `GeneFilterByProteinCoding`.

The input consists of two groups of RNA data analysis from tissues under different conditions: control/altered state.
For example, a disease-affected group and a control group.

Unlike the dichotomous case, where groups are divided only into "control/altered state", additional conditions exist under which the experiments were conducted. For instance, multiple RNA plates were used, or tissues originated from different parts of the body.
RNA data must follow "normal distribution", which is typical for RNA microarray analysis.

For RNA data following a "normal distribution", refer to `LMMDGEDream`, `LMMDGEJulia`, or `LMMDGELmer`.



```{r setup, include=FALSE}
library(DrugRepurposing)
options(warn = -1)
```
```{r}
library(DrugRepurposing)

sample_id <- c("GSM1", "GSM2", "GSM3", "GSM4", "GSM5", "GSM6",
               "GSM7", "GSM8", "GSM9", "GSM10", "GSM11", "GSM12")
rna_seq_data <-
  matrix(
    # rna seq data analysis
    c(632, 944, 219, 547, 895, 610, 369, 427, 763, 3, 772, 980,
      51, 659, 275, 911, 449, 509, 742, 410, 935, 66, 238, 358,
      323, 681, 521, 947, 133, 492, 459, 224, 876, 692, 927, 836),
    nrow = 3,
    ncol = 12,
    byrow = TRUE,
    dimnames = list(
      # gene IDs
      c(203228, 6647, 2521),
      sample_id
    )
  )
# the first column represents the row names (gene IDs)
# rna_seq_data =
#          GSM1   GSM2   GSM3   GSM4   GSM5 ...
# 203228   632    944    219    547    895  ...
# 6647     51     659    275    911    449  ...
# 2521     323    681    521    947    133  ...

tissue_status <- c("ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS", "ALS",
                   "ALS", "ALS", "Control", "Control")
# The first level must be "Control" state
tissue_status_factor <- factor(tissue_status, levels = c("Control", "ALS"))
rna_seq_metadata <- data.frame(
  sample_id = sample_id,
  tissue = c("Spinal Cord Lumbar", "Spinal Cord Lumbar", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Cerebellum", "Cerebellum", "Cerebellum",
             "Cerebellum", "Cerebellum", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Cerebellum"),
  tissue_status = tissue_status_factor
)

# rna_seq_metadata =
# sample_id   tissue                  subject_id     tissue_status
# GSM1        Spinal Cord Lumbar      NEUEL133AK6    ALS Spectrum MND
# GSM2        Spinal Cord Lumbar      NEUYV496XLP    ALS Spectrum MND
# GSM3        Spinal Cord Lumbar      NEUZU200WEQ    ALS Spectrum MND
# GSM4        Spinal Cord Lumbar      NEUEC006FND    Non-Neurological Control
# GSM5        Spinal Cord Lumbar      NEUKM699KKH    ALS Spectrum MND
# GSM6        Cerebellum              NEUJZ255ZC5    ALS Spectrum MND
# GSM7        Cerebellum              NEUJZ233ZL3    Non-Neurological Control

formula <- ~tissue_status + (1 | tissue)

lmmVoomDGELmer <- LMMVoomDGELmer$new()
lmmVoomDGELmer$compute(
  rna_seq_data = rna_seq_data,
  rna_seq_metadata = rna_seq_metadata,
  formula = formula,
  filter_by_protein_coding = FALSE
)
```
