---
title: "LMM Voom Drug Repurpose"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
 %\VignetteIndexEntry{LMM Voom Drug Repurpose}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
# Description
Given a disease RNA-Seq data analysis and a list of drug DGEs, `LMMVoomDrugRepurpose` computes the
connectivity score between the disease and the drugs.

`LMMVoomDrugRepurpose` calculates differential gene expression from RNA-seq data using a
Linear Mixed Model, and then computes the connectivity score between the disease DGE and the drug DGEs.

RNA data must follow a negative binomial distribution, typically data from RNA-Seq-like analyses.

This class creates an instance of `LMMDrugRepurposeCore` and uses it for the computation.
```{r setup, include=FALSE}
library(DrugRepurposing)
options(warn = -1)
```
```{r}
library(DrugRepurposing)

rna_seq_data <- data.frame(
  GSM1 = c(262, 281897, 463, 12301, 284, 2966, 4842, 9966),
  GSM2 = c(116, 337026, 261, 3277, 132, 2006, 3699, 12406),
  GSM3 = c(111, 111736, 193, 862, 88, 1371, 2536, 1531),
  GSM4 = c(162, 219471, 281, 5990, 280, 2241, 4527, 12941),
  GSM5 = c(82, 471, 981, 590, 380, 241, 427, 941),
  GSM6 = c(92, 9471, 201, 1990, 80, 41, 1527, 2941),
  GSM7 = c(192, 2191, 181, 3990, 480, 1241, 7527, 1941)
)
rownames(rna_seq_data) <- c(1, 2, 9, 12, 13, 14, 21, 25) # gene id

# rna_seq_data =
#      GSM1     GSM2      GSM3     GSM4     GSM5   GSM6   GSM7
# 2    262      116       111      162      82     92     192
# 2    281897   337026    111736   219471   471    9471   2191
# 3    463      261       193      281      981    201    181
# 9    12301    3277      862      5990     590    1990   3990
# 10   284      132       88       280      380    80     480
# 11   2966     2006      1371     2241     241    41     1241
# 12   4842     3699      2536     4527     427    1527   7527
# 13   9966     12406     1531     12941    941    2941   1941

tissue_status <- c("ALS Spectrum MND", "ALS Spectrum MND", "ALS Spectrum MND",
                   "Non-Neurological Control", "ALS Spectrum MND", "ALS Spectrum MND",
                   "Non-Neurological Control")
# The first level must be "Control" state
tissue_status_factor <- factor(tissue_status, levels = c("Non-Neurological Control", "ALS Spectrum MND"))

rna_seq_metadata <- data.frame(
  sample_id = c("GSM1", "GSM2", "GSM3", "GSM4", "GSM5", "GSM6", "GSM7"),
  tissue = c("Spinal Cord Lumbar", "Spinal Cord Lumbar", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Spinal Cord Lumbar", "Cerebellum", "Cerebellum"),
  subject_id = c("NEUEL133AK6", "NEUYV496XLP", "NEUZU200WEQ", "NEUEC006FND",
                 "NEUKM699KKH", "NEUJZ255ZC5", "NEUJZ233ZL3"),
  tissue_status = tissue_status_factor
)

# rna_seq_metadata =
# sample_id   tissue                  subject_id     tissue_status
# GSM1        Spinal Cord Lumbar      NEUEL133AK6    ALS Spectrum MND
# GSM2        Spinal Cord Lumbar      NEUYV496XLP    ALS Spectrum MND
# GSM3        Spinal Cord Lumbar      NEUZU200WEQ    ALS Spectrum MND
# GSM4        Spinal Cord Lumbar      NEUEC006FND    Non-Neurological Control
# GSM5        Spinal Cord Lumbar      NEUKM699KKH    ALS Spectrum MND
# GSM6        Cerebellum              NEUJZ255ZC5    ALS Spectrum MND
# GSM7        Cerebellum              NEUJZ233ZL3    Non-Neurological Control

formula <- ~tissue_status + (1 | tissue)


drugs_genes <- c(1, 2, 9, 12, 13, 14, 21, 25)

drugs <- data.frame(
  name = c("A-23187", "A-443644", "AG-490", "AG-494"),
  filename = c("A-23187", "A-443644", "AG-490", "AG-494")
)

A_23187_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 31),
  estimate = c(1.98564, -1.51579, 0.33099, -0.935718, -3.818529,
               1.848352, -4.332752, 3.591474, -0.345235423)
)

A_443644_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 6158),
  estimate = c(-1.55774, -4.4155, 1.21015, 4.8705936, -0.1259465,
               -2.6003531, -0.3530122, -3.5426363, -3.252352)
)

AG_490_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 37),
  estimate = c(-1.35143, 0.58724, -1.31592, 0.22267776, -0.45726445,
               -2.80993289, -0.05550198, 1.76931022, 0.1423542)
)

AG_494_drug <- data.frame(
  gene_id = c(1, 2, 9, 12, 13, 14, 21, 25, 6158),
  estimate = c(-3.43357, -0.14896, 0.90733, 3.084164, -2.594041,
               4.304979, -4.944015, -2.075178, 1.4523542)
)

drugs_rna_data <- list()
drugs_rna_data[["A-23187"]] <- A_23187_drug
drugs_rna_data[["A-443644"]] <- A_443644_drug
drugs_rna_data[["AG-490"]] <- AG_490_drug
drugs_rna_data[["AG-494"]] <- AG_494_drug


EasyDrugSignatureLoader <- R6Class(
  "EasyDrugSignatureLoader",
  inherit = DrugSignatureLoaderAbstract,
  public = list(
    initialize = function(drugs_rna_data) {
      private$drugs_rna_data <- drugs_rna_data
    },
    load = function(drug_name) {
      return(private$drugs_rna_data[[drug_name]])
    }
  ),
  private = list(
    drugs_rna_data = NA
  )
)

easyDrugSignatureLoader <- EasyDrugSignatureLoader$new(drugs_rna_data)

lmmVoomDGEJulia <- LMMVoomDGEDream$new()
lmmVoomDrugRepurpose <- LMMVoomDrugRepurpose$new(lmmVoomDGE = lmmVoomDGEJulia, drugSignatureLoader = easyDrugSignatureLoader)
lmmVoomDrugRepurpose$compute(
  rna_seq_data = rna_seq_data,
  rna_seq_metadata = rna_seq_metadata,
  formula = formula,
  drugs = drugs,
  drugs_genes = drugs_genes,
  random_distribution_size = 10, # 10 is a poor value used only to speed up
                                 # computation; an appropriate value is 10^5
  disease_name = "ALS",
  parallel_computation = FALSE,
  filter_by_protein_coding = FALSE,
  signature_mapper_parameter = 8
)
```