---
title: "Linear Mixed Model Voom DGE by Dream With eBayes"
output: rmarkdown::html_vignette
author: "Carmelo Catalano"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
vignette: >
 %\VignetteIndexEntry{Linear Mixed Model Voom DGE by Dream With eBayes}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
# Description
Derived class of `LMMVoomDGEAbstract`.

Computes the Voom Linear Mixed Model (LMM) differential gene expression (DGE) analysis using the `dream` function from the `variancePartition` package.

After the `dream` computation is completed "Empirical Bayes Statistics for Differential Expression" are applied (`variancePartition::eBayes`).

It relies on three collaborators: `LMMVoom`, `DGEMapper`, and `GeneFilterByProteinCoding`.


The input consists of two groups of RNA data analysis from tissues under different conditions: control/altered state.
For example, a disease-affected group and a control group.

Unlike the dichotomous case, where groups are divided only into "control/altered state", additional conditions exist under which the experiments were conducted. For instance, multiple RNA plates were used, or tissues originated from different parts of the body.
RNA data must follow "normal distribution", which is typical for RNA microarray analysis.

For RNA data following a "normal distribution", refer to `LMMDGEDream`, `LMMDGEDreamWithEBayes`, `LMMDGEJulia`, or `LMMDGELmer`.



```{r setup, include=FALSE}
library(DrugRepurposing)
options(warn = -1)
```
```{r}
library(DrugRepurposing)

rna_seq_data <- data.frame(
  GSM1 = c(262, 281897, 463, 12301, 284, 2966, 4842, 9966),
  GSM2 = c(116, 337026, 261, 3277, 132, 2006, 3699, 12406),
  GSM3 = c(111, 111736, 193, 862, 88, 1371, 2536, 1531),
  GSM4 = c(162, 219471, 281, 5990, 280, 2241, 4527, 12941),
  GSM5 = c(82, 471, 981, 590, 380, 241, 427, 941),
  GSM6 = c(92, 9471, 201, 1990, 80, 41, 1527, 2941),
  GSM7 = c(192, 2191, 181, 3990, 480, 1241, 7527, 1941)
)
rownames(rna_seq_data) <- c(1, 2, 9, 12, 13, 14, 21, 25) # gene id

# rna_seq_data =
#      GSM1     GSM2      GSM3     GSM4     GSM5   GSM6   GSM7
# 2    262      116       111      162      82     92     192
# 2    281897   337026    111736   219471   471    9471   2191
# 3    463      261       193      281      981    201    181
# 9    12301    3277      862      5990     590    1990   3990
# 10   284      132       88       280      380    80     480
# 11   2966     2006      1371     2241     241    41     1241
# 12   4842     3699      2536     4527     427    1527   7527
# 13   9966     12406     1531     12941    941    2941   1941
tissue_status <- c("ALS Spectrum MND", "ALS Spectrum MND", "ALS Spectrum MND",
                   "Non-Neurological Control", "ALS Spectrum MND", "ALS Spectrum MND",
                   "Non-Neurological Control")
# The first level must be "Control" state
tissue_status_factor <- factor(tissue_status, levels = c("Non-Neurological Control", "ALS Spectrum MND"))

rna_seq_metadata <- data.frame(
  sample_id = c("GSM1", "GSM2", "GSM3", "GSM4", "GSM5", "GSM6", "GSM7"),
  tissue = c("Spinal Cord Lumbar", "Spinal Cord Lumbar", "Spinal Cord Lumbar",
             "Spinal Cord Lumbar", "Spinal Cord Lumbar", "Cerebellum", "Cerebellum"),
  subject_id = c("NEUEL133AK6", "NEUYV496XLP", "NEUZU200WEQ", "NEUEC006FND",
                 "NEUKM699KKH", "NEUJZ255ZC5", "NEUJZ233ZL3"),
  tissue_status = tissue_status_factor
)

# rna_seq_metadata =
# sample_id   tissue                  subject_id     tissue_status
# GSM1        Spinal Cord Lumbar      NEUEL133AK6    ALS Spectrum MND
# GSM2        Spinal Cord Lumbar      NEUYV496XLP    ALS Spectrum MND
# GSM3        Spinal Cord Lumbar      NEUZU200WEQ    ALS Spectrum MND
# GSM4        Spinal Cord Lumbar      NEUEC006FND    Non-Neurological Control
# GSM5        Spinal Cord Lumbar      NEUKM699KKH    ALS Spectrum MND
# GSM6        Cerebellum              NEUJZ255ZC5    ALS Spectrum MND
# GSM7        Cerebellum              NEUJZ233ZL3    Non-Neurological Control

formula <- ~tissue_status + (1 | tissue)

lmmVoomDGEDreamWithEBayes <- LMMVoomDGEDreamWithEBayes$new()
lmmVoomDGEDreamWithEBayes$compute(
  rna_seq_data = rna_seq_data,
  rna_seq_metadata = rna_seq_metadata,
  formula = formula,
  filter_by_protein_coding = FALSE
)
```
